---
title: "Data preparation Muraro (2016)"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
bibliography: ../ref.bibtex
---

```{r loading packages}
suppressMessages(library(scRNAseq))
suppressMessages(library(AnnotationHub))
suppressMessages(library(scater))
suppressMessages(library(scran))
suppressMessages(library(scuttle))
suppressMessages(library(scDblFinder))
suppressMessages(library(BiocParallel))
suppressMessages(library(knitr))
suppressMessages(library(kableExtra))
```

# Loading the data

We loaded the Muraro dataset directly from the _scRNAseq_ package as a SingleCellExperiment object. The dataset contains CEL-seq scRNA-seq data from human pancreas cells from 4 different donors [@muraro2016single].

```{r loading data, warning=FALSE, message=FALSE}
sce.muraro <- suppressMessages(MuraroPancreasData(ensembl = TRUE))
dim(sce.muraro)
```

The option __ensebml = TRUE__ ensures that the gene symbols are converted to Ensembl IDs. Any genes that could not be matched to an Ensembl ID were discarded (2110 genes). We were left with $16940$ genes from $3072$ different cells. 

# Quality Control

Quality control was performed on the loaded Baron dataset. 

First we tried to identify cells with relatively high proportion of mitochondrial genes, which should be removed, as this indicates damaged nuclei, and thus low-quality cells.

## Visual Inspection before QC
```{r mitochondrial genes}
location <- rowRanges(sce.muraro)
isMito <- grepl("^mt-", location$originalName, ignore.case=TRUE) | 
  seqnames(location)=="MT" | seqnames(location)=="mt"
ifelse(any(isMito), "There has been found at least one mitochondrial gene",
       "There have not been found any mitochondrial genes")
df <- perCellQCMetrics(sce.muraro)
```

Suprisingly, we saw that there are no mitochondrial genes left in the dataset. We assume that the authors have removed all mitochondrial genes beforehand.

```{r plot1, warning=FALSE, message=FALSE}
#get QC metrics:
sce.muraro <- addPerCellQC(sce.muraro)
sce.muraro <- addPerFeatureQC(sce.muraro)

qc <- as.data.frame(colData(sce.muraro))
ggplot(qc, aes(sum)) + geom_histogram() + facet_wrap(~donor) + scale_x_continuous(trans = 'log10') + ggtitle("Libary sizes of every genes split by donor") + ylab("Number of genes") + xlab("Library Size")
```

The histograms of the library sizes show bimodal distributions, especially in donor D28. What we would like to have though, are unimodal, approximately normally distributed library sizes.

```{r plot2}
ggplot(qc, aes(sum, detected)) + geom_point() + geom_density2d() + ylab("Number of detected genes") + xlab("Library size") + scale_x_continuous(trans = 'log10') + scale_y_continuous(trans = 'log10') + ggtitle("Library size vs. number of detected genes per cell")
```

As we can see from the plot, the range of library sizes as well as the range of the number of detected genes is too large. 

## QC

Thus, we have to perform quantitative outlier detection, in order to discard cells that have very low library size or a too low number of detected genes. Additionaly, we want to discard cells with a too high ERCC spike-in content, as this is symptomatic of loss of endogenous RNA [@amezquita2020orchestrating].

```{r, warning=FALSE, message=FALSE}
reasons <- quickPerCellQC(df, 
    sub.fields="altexps_ERCC_percent")
discard_reasons <- as.data.frame(colSums(as.matrix(reasons)))
names(discard_reasons) <- "counts"
knitr::kable(discard_reasons) %>%
  kable_styling()%>%
  row_spec(4,bold=T)
#spike-ins
sce.muraro$exclude_ERCC <- reasons$high_altexps_ERCC_percent
sce.muraro$discard <- reasons$discard
plotColData(sce.muraro, x="donor", y="altexps_ERCC_percent", colour_by = I(reasons$high_altexps_ERCC_percent)) + xlab("Donor") + 
  ggtitle("Remove outliers based on ERCC spike-ins")

# remove outliers
sce.muraro <- sce.muraro[,!apply(reasons, MARGIN = 1, any)]
```

From the table, we see that a total of $726$ cells were excluded. Further, we see that most cells were excluded for more than one reason. In total, $611$ cells were excluded due to low library size, $669$ are exluded due to low number of features and $696$ cells are excluded based on high spike-in counts. 

```{r qc, warning=FALSE, message=FALSE}
# get QC metrics:
sce.muraro <- addPerCellQC(sce.muraro)
sce.muraro <- addPerFeatureQC(sce.muraro)
```

## Visual inspection after QC

```{r qc visual, warning=FALSE, message=FALSE}
# plots
qc <- as.data.frame(colData(sce.muraro))

ggplot(qc, aes(sum)) + geom_histogram() + facet_wrap(~donor) + ggtitle("Libary sizes of every genes split by donor") + scale_x_continuous(trans = "log10") + ylab("Number of genes") + xlab("log10(Library Size)")
```

Looking at the histogram, we see that the library sizes are now reasonably distributed. Additionally, we see that most cells that have been excluded, came from donor D28. 

```{r plot3, warning=FALSE, message=FALSE}
ggplot(qc, aes(sum, detected)) + geom_point() + geom_density2d() + ylab("Number of detected genes") + xlab("Library size)") + scale_x_continuous(trans = 'log10') + scale_y_continuous(trans = 'log10') + ggtitle("Library size vs. number of detected genes per cell")
```

After removal of all outliers, we see that the range of library sizes, as well as the number of detected genes, is now much narrower than before. 
  

# Unifying Label Names

In order to be able to compare the two datasets by @muraro2016single and @baron2016single, we need to ensure that they have the same cell type labels. Therefore we rename "duct" to "ductal and "pp" to "gamma", similarly performed by @lin2019scclassify. Further, we remove all cells that are either labelled "unclear" or "mesenchymal", because these label types are not present in the Baron dataset. We are left with 2022 cells, belonging to 8 different cell types.  


```{r harmonize labels, warning=FALSE, message=FALSE}
sce.muraro$label[sce.muraro$label=="duct"] <- "ductal" #renaming
sce.muraro$label[sce.muraro$label=="pp"] <- "gamma" #renaming
sce.muraro <- sce.muraro[,!(sce.muraro$label %in% c(NA,"unclear", "mesenchymal"))] #removing
muraro.labels.counts <- as.data.frame(table(sce.muraro$label))
names(muraro.labels.counts) <- c("cell type", "frequency")
total <- data.frame("total", sum(muraro.labels.counts$frequency))
names(total) <- c("cell type", "frequency")
muraro.labels.counts <- rbind(muraro.labels.counts, total)
saveRDS(muraro.labels.counts, file="data/muraro.labels.counts.rds")
knitr::kable(muraro.labels.counts) %>%
  kable_styling()%>%
  row_spec(9,bold=T)
```

# tSNE plot

Finally, we produce normalized counts and perform dimensionality reduction for visual inspection purposes. 
```{r log counts, warning=FALSE, message=FALSE}
sce.muraro <- logNormCounts(sce.muraro)
```

```{r PCA and t-SNE, warning=FALSE, message=FALSE}
sce.muraro <- runPCA(sce.muraro)
sce.muraro <- runTSNE(sce.muraro, dimred="PCA")
```

```{r}
plotReducedDim(sce.muraro, "TSNE", colour_by="label") #original labels
```
We can see that the "true" cell type labels correspond quite nicely to the clusters in the tSNE plot. 

## Save object

```{r save sce object, warning=FALSE, message=FALSE}
saveRDS(sce.muraro,file = "data/sce_muraro.rds")
```

# References

